---
title: "Wahlund Effect"
author: "Jinliang Yang"
date: "2025-02-11"
output:
  html_document: default
  pdf_document: default
---

# Population subdivision


| Genotype | $A_1A_1$   | $A_1A_2$    | $A_2A_2$   
|:-------:|: ------ :|:-------:|:-------:|   
| sub-pop1  | 4    | 32    | 64    |   
| sub-pop2  | 16     | 48    | 36    |   
| Total     | 20    | 80    | 100    |   


The frequencies of the $A_1$ allele in sub-pop1 and sub-pop2:
\begin{align*}
p_1 & = \frac{2 \times 4 + 32}{2 \times 100} = 0.2 \\
p_2 & = \frac{2 \times 16 + 48}{2 \times 100} = 0.4 \\
\end{align*}


The frequency of the $A_1$ allele in the combined population:

\begin{align*}
\bar{p} & = \frac{2 \times 20 + 80}{2 \times 200} = 0.3 \\
\end{align*}


The expected value of heterozygotes for the combined population is:

\begin{align*}
2\bar{p}\bar{q} \times n = 2 \times 0.3 \times 0.7 \times 200 = 84 \\ 
\end{align*}

---

# Wahlund Effect

The _perceived deficiency_ of heterozygotes due to treating two different populations as one --- the __Wahlund effect__. 

```{r}
# get heterozygote genotype freq based on HWE
geth <- function(p){
  return(2*p*(1-p))
}

# for a give allele freq, calculate the heterozygote genotype freq
p <- seq(0, 1, by=0.01)
plot(p, geth(p), type="l", lwd=3, col="red", xlab="Allele freq (p)", ylab="Heterozygosity (h)")
```
# for a sub-population 1

```{r}
p <- seq(0, 1, by=0.01)
plot(p, geth(p), type="l", lwd=3, col="red", xlab="Allele freq (p)", ylab="Heterozygosity (H)")
lines(c(0.2, 0.2), c(0, geth(p=0.2)), lty=2, lwd=3) # add a line segment
lines(c(0, 0.2), c(geth(p=0.2), geth(p=0.2)), lty=2, lwd=3)
```
# for a sub-population 2

```{r}
plot(p, geth(p), type="l", lwd=3, col="red", xlab="Allele freq (p)", ylab="Heterozygosity (H)")
lines(c(0.2, 0.2), c(0, geth(p=0.2)), lty=2, lwd=3) # add a line segment
lines(c(0, 0.2), c(geth(p=0.2), geth(p=0.2)), lty=2, lwd=3)
lines(c(0.4, 0.4), c(0, geth(p=0.4)), lty=2, lwd=3) # add a line segment
lines(c(0, 0.4), c(geth(p=0.4), geth(p=0.4)), lty=2, lwd=3)
```


```{r wahlund, echo=TRUE, fig.align='center', fig.height=6}
geth <- function(p){return(2*p*(1-p))}
p <- seq(0, 1, by=0.01)
plot(p, geth(p), type="l", lwd=3, col="red", xlab="Allele freq (p)", ylab="Heterozygosity (H)")
lines(c(0.2, 0.2), c(0, geth(p=0.2)), lty=2, lwd=3) # add a line segment
lines(c(0, 0.2), c(geth(p=0.2), geth(p=0.2)), lty=2, lwd=3)
lines(c(0.4, 0.4), c(0, geth(p=0.4)), lty=2, lwd=3) # add a line segment
lines(c(0, 0.4), c(geth(p=0.4), geth(p=0.4)), lty=2, lwd=3)
lines(c(0.3, 0.3), c(0, geth(p=0.3)), lty=2, lwd=3) 
lines(c(0, 0.3), c(geth(p=0.3), geth(p=0.3)), lty=2, lwd=3, col="blue")
lines(c(0.2, 0.4), c(geth(p=0.2), geth(p=0.4)), lty=1, lwd=3, col="blue")
lines(c(0, 0.3), c(mean(c(geth(p=0.2), geth(p=0.4))), mean(c(geth(p=0.2), geth(p=0.4)))), lty=2, lwd=3, col="red")
```

---

# Wahlund Effect



- The blue dashed horizontal segment is the $H_{exp} = 2 \bar{p}\bar{q}$

- The red dashed horizontal segment is the $H_{obs}$ 
  - mean value of the $H_1$ and $H_2$

- Because the curve is concave downward, $H_{obs}$ is always less than $H_{exp}$


Even if subpopulations are __partially isolated__, the Wahlund effect is held true.

