---
title: "Angsd and XP-CLR"
author: "Jinliang Yang"
date: "2025-03-13"
output:
  html_document:
    df_print: paged
  word_document: default
---

## Normalize the path:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
knitr::opts_knit$set(root.dir=normalizePath('../../')) 
```


# A procedure to detect linked selection

#### Request a computation node

```{bash,eval=F}
srun --qos=short --nodes=1 --licenses=common --ntasks=1 --mem 6G --licenses=common --time 2:00:00 --pty bash
```

### Single read sampling approach for PCA---Corrects for unequal sequencing depth bias

```{bash,eval=F}
module load angsd/0.937
angsd -bam bam_list.txt -minMapQ 30 -minQ 20 -GL 2  -doMajorMinor 1 -doMaf 1 -SNP_pval 2e-6 -doIBS 1 -doCounts 1 -doCov 1 -makeMatrix 1 -minMaf 0.05 -out sorghum -P 4
# then copy results to cache
```

- plot PCA

```{r}
m <- as.matrix(read.table("cache/sorghum.covMat"))
eig <- eigen(m)
write.table(eig$vectors, "cache/eigenvectors.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
write.table(eig$values, "cache/eigenvalues.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
plot(eig$vectors[,1:2], lwd=2, xlab="PC 1",ylab="PC 2", main="Principal components", col="darkgreen",pch=16,las=1,cex=2)
```

## Powerful tools for population parameter calculation: [**Plink**](https://www.cog-genomics.org/plink/1.9/) and [**vcftools**](https://vcftools.github.io/man_latest.html)

### Functions in plink
- Prepare binary files for plink
```{bash,eval=F}
cd ../02vcf_files/
gunzip -c TEO_LR_MZ_test.vcf.gz > TEO_LR_MZ_test.vcf
module load plink
plink --vcf TEO_LR_MZ_test.vcf --make-bed --out  TEO_LR_MZ_test
```
- Minor allele frequency
```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --freq --out TEO_LR_MZ_test
```
- Missing data
```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --missing --out TEO_LR_MZ_test
```
- Hardy-Weinberg equilibrium
```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --hardy --out TEO_LR_MZ_test
```
- Homozygous genotype counts for each sample
```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --het --out TEO_LR_MZ_test
```
- multiple parameters
```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --freq --missing --hardy --het --out TEO_LR_MZ_test
```

- PCA
```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --pca 10 --out TEO_LR_MZ_test
```

- Variant pruning

```{bash,eval=F}
plink --bfile TEO_LR_MZ_test --indep-pairwise 100 10 0.1 --out outfile
plink --bfile TEO_LR_MZ_test --extract outfile.prune.in --make-bed --out prunedData
```

- bfiles to vcf
```{bash,eval=F}
plink --bfile prunedData --recode vcf --out prunedData
```

### Functions in Vcftools
- Allele frequency from Chr5
```{bash,eval=F}
vcftools --gzvcf TEO_LR_MZ_test.vcf.gz --freq --chr 10 --out chr10_analysis
```
- Filtering according SNP missing rate or sample name
```{bash,eval=F}
vcftools --vcf TEO_LR_MZ_test.vcf --max-missing 0.7 --recode --out TEO_LR_MZ_test_miss03

vcftools --vcf TEO_LR_MZ_test.vcf --keep maize.txt --recode --out maize
```

- Nucleotide diversity
```{bash,eval=F}
vcftools --vcf maize.recode.vcf --window-pi 25000 --window-pi-step 5000 --out MZ_test
```
- Fst estimate from Weir and Cockerham’s 1984 paper
```{bash,eval=F}
vcftools --vcf TEO_LR_MZ_test.vcf --weir-fst-pop teo.txt --weir-fst-pop maize.txt --fst-window-size 25000 --fst-window-step 5000 --out TEO_vs_MZ_w25k
```
- Small sample size and low allele frequency can casue negative fst, we can set them to 0
```{bash,eval=F}
awk '{if($5 < 0) $5=0; print}' TEO_vs_MZ_w25k.windowed.weir.fst > TEO_vs_MZ_w25k.windowed_fixed.fst
```

### Selection scan using [**XP-CLR**](https://github.com/hardingnj/xpclr)

The original paper is [**here**](https://pubmed.ncbi.nlm.nih.gov/20086244/)

```{bash,eval=F}
module load xpclr/1.1 bcftools/1.21

bcftools view -Ov -o merged.vcf merged.bcf

xpclr --out xpclr_results --input merged.vcf --format vcf --samplesA maize.txt --samplesB teo.txt --chr 10 --ld 0.7 --start 98000000 --stop 105000000 --maxsnps 200 --minsnps 200 --size 25000 --step 5000
```
- Object is pop A. Reference is pop B
- --out full path is needed




